---
title: "GeOnG Workshop"
author: "Linda Menk"
date: "2022-10-14"
output: 
  html_document:
    theme: yeti
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
      number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Malaria risk assessment for Burundi



```{r loading libraries, echo = FALSE, message = FALSE}

#Data access
library(malariaAtlas) # Malaria Atlas Project
library(rnaturalearth) # NaturalEarth
library(wpgpDownloadR) # Worldpop

#Mapping
library(tmap) # To create interactive maps easily
library(RColorBrewer) # To access popular color palettes for our maps

#Geoprocessing
library(sf) # To work with spatial vector data
library(terra) # To work with raster data
library(raster) # To work with raster data (predecessor of terra)
library(magrittr) # For data tidying and manupulation (not limited to spatial data)
```
# Part 1: Data access  

For our malaria risk assessment, we will use data from the [Malaria Atlas Project](https://malariaatlas.org/) and [Worldpop](https://www.worldpop.org/). 


## Administrative boundaries of our area of interest (AOI)
```{r admin, message = FALSE}
# Downloads the geometric shape of our AOI on admin 2 level from the Malaria Atlas Project. 
AOI_shp <- malariaAtlas::getShp(ISO = "BDI", admin_level = "admin2")  

# Creates a simple plot of the data that we have just downloaded
plot(AOI_shp)

```


## Malaria incidence  

```{r malaria incidence, message = FALSE}

# Downloads malaria indicator data for our AOI from Malaria Atlas Project
AOI_Pf2020 <- getRaster(surface = "Plasmodium  falciparum Incidence version 2020", shp = AOI_shp)

# Makes a basic plot of the downloaded malaria indicator data.
plot(AOI_Pf2020)

```

## Accessibility to healthcare services
```{r access healthcare, message = FALSE}

# Downloads data for estimated walking time/minutes to the nearest healthcare facility for our AOI from Amalaria Atlas project
AOI_access <- getRaster(surface = "Walking-only travel time to healthcare map without access to motorized transport", shp = AOI_shp)

# Makes a basic plot of the downloaded accessibility indicator data
plot(AOI_access)

```
# Population statistics  
This dataset was produced based on the 2020 population census/projection-based estimates for 2020.[More Info.](https://hub.worldpop.org/geodata/summary?id=49625)
```{r Worldpop, message = FALSE, warning = FALSE}

# Downloads the population data (covariate) and saves it to disk
pop_path <- wpgpGetCountryDataset(ISO3 = "BDI",
                             covariate = "ppp_2020",
                             destDir = tempdir())

# Uses the "raster" function to load the downloaded data from disk into R
pop <- raster(pop_path)

# The downloaded data has a spatial resolution of 100m. For our analysis we do not need such fine data, it only slows our analysis down.
# Therefore, we aggregate the data by the factor 10 to get a spatial resolution of 1km. 
AOI_pop <- aggregate(pop, fact = 10, fun = sum)

# Clears the "pop" data from memory. We don't need it anymore as we continue working with the aggregated data.
remove(pop)

# Makes a basic plot of the population indicator data
plot(AOI_pop)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
# Part 2: Data manipulation and analysis 
# Turn the raw data into indicators  
```{r analysis, message = FALSE}
#----------------
# Analysis - Count Pop, Incidence, Avg. dist to healthcare per Admin area

# Convert administrative borders to a sf object
AOI_sf <- st_as_sf(AOI_shp)

# Sums the population per district
pop_sum <- terra::extract(AOI_pop, AOI_sf, fun = sum, ID = TRUE, na.rm = TRUE) %>% 
  round(digits =2)

pop_district <- cbind(pop_sum,AOI_sf)

# Mean plasmodium falciparum indicence per district
Pf20_mean <- terra::extract(AOI_Pf2020, AOI_sf, fun = mean, ID = TRUE, na.rm = TRUE) %>% 
  round(digits = 2)

Pf20_district <- cbind(Pf20_mean, AOI_sf)

# Mean distance to the nearest healthcare facility per district
access_mean <- terra::extract(AOI_access, AOI_sf, fun = mean, ID = TRUE, na.rm = TRUE) %>% 
  round(digits = 2)

access_district <- cbind(access_mean, AOI_sf)


# Bind all the indicators to the same geospatial outlines
AOI_district <- cbind(pop_district, Pf20_district, access_district)

# Removes the Lake Tanganyika district because it covers only lake areas.
AOI_district <- AOI_district[!(AOI_district$name_2=="Lake Tanganyika"),]

remove(pop_sum, Pf20_mean, access_mean)
```

# Create composite indicator
```{r created composite indicator, message = FALSE}

# Perform min-max normalization on population indicator
AOI_district$pop_norm <- (AOI_district$pop_sum - min(AOI_district$pop_sum))/(max(AOI_district$pop_sum) - min(AOI_district$pop_sum))

# Perform min-max normalization on malaria incidence indicator
AOI_district$pf_norm <- (AOI_district$Pf20_mean - min(AOI_district$Pf20_mean))/(max(AOI_district$Pf20_mean) - min(AOI_district$Pf20_mean))

# Perform min-max normalization on accessibility indicator
AOI_district$access_norm <- (AOI_district$access_mean - min(AOI_district$access_mean))/(max(AOI_district$access_mean) - min(AOI_district$access_mean))

# Sum all three sub-indicators 
AOI_district$composite_indicator <- (AOI_district$pop_norm + AOI_district$pf_norm + AOI_district$access_norm)/3  
AOI_district$composite_indicator <- round(AOI_district$composite_indicator, digits = 2)

```

# Part 3: Visualizing results as maps
```{r map, warning = FALSE, error = FALSE, message = FALSE, fig.height= 12}

tmap_mode("view")

pal <- c('#60b564','#fed400')

pop <- tm_shape(AOI_district) +
  tm_polygons(col = "pop_sum", palette = pal, popup.vars = c("Pop"="pop_sum")) +
  tm_borders(lwd = 0.2) + 
  tm_layout(title = "Total Population") + 
  tm_fill(col = '#BFE1F4') + 
  tm_borders(lwd = 1, col = "black") 

access <- tm_shape(AOI_district) +
  tm_polygons(col = "access_mean", palette = pal, id = "name_2", popup.vars = c("Access"="access_mean")) +
  tm_borders(lwd = 0.2) + 
  tm_layout(title = "Healthcare accessibility") + 
  tm_fill(col = '#BFE1F4') + 
  tm_borders(lwd = 1, col = "black") 

pf20 <- tm_shape(AOI_district) +
  tm_polygons(col = "Pf20_mean", palette = pal, id = "name_2", popup.vars = c("Malaria incidence"="Pf20_mean")) +
  tm_borders(lwd = 0.2) + 
  tm_layout(title = "Malaria incidence") +
  tm_fill(col = '#BFE1F4') + 
  tm_borders(lwd = 1, col = "black")

# Visualize composite indicator as a map
risk <- tm_shape(AOI_district) +
  tm_polygons(col = "composite_indicator", palette = pal, id = "composite_indicator", popup.vars = c("Name" = "name_2", "Risk index"="composite_indicator", "Population" = "pop_norm", "Accessibility" = "access_norm", "Malaria indcidence" = "pf_norm")) +
  tm_borders(lwd = 0.2) + 
  tm_layout(title = "Composite risk indicator") + 
  tm_fill(col = '#BFE1F4') + 
  tm_borders(lwd = 1, col = "black") 

x <- list(pop, access, pf20, risk)

tmap_arrange(x, sync = TRUE)

```